Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-06-29 submission/manuscript.Rmd
intro draft

---
title:  Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C
output: pdf_document
geometry: margin = 0.5in
font-size: 11pt
header-includes:
   - \usepackage{caption}
   - \captionsetup[figure]{font=footnotesize, labelfont=bf}
   - \usepackage{sectsty} 
   - \usepackage{helvet}
   - \renewcommand{\familydefault}{\sfdefault}
   - \usepackage{lineno}
   - \linenumbers
   - \renewcommand\linenumberfont{\normalfont\small}
   - \usepackage{setspace}
   - \spacing{1.5}
   - \pagenumbering{arabic}

@@

@@

@@

RNA sequencing (RNA-seq) studies surprisingly revealed the aberrant expression of testis genes within the adult _Kdm5c_ knockout (-KO) hippocampus[@scandagliaLossKdm5cCauses2017]. Given the high abundance of testis-enriched genes within the mouse transcriptome, it is currently unclear if their misexpression in the _Kdm5c_-KO brain is indicative of random de-repression throughout the genome or instead reflects an impairment in tissue identity. To distinguish these possibilities, we globally assessed the expression of previously characterized tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017] in our previously published mRNA-seq datasets of the _Kdm5c_-KO amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020]

@@

# Discussion
* tissue-biased gene expression:
	* However unlike the gonadal-biased DEGs, many liver and muscle-biased DEGs have a known involvement in brain function. For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory (Figure 1E).
* Otx2 is properly expressed in EpiLCs and prevents pgc identity https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/

>>>>

---
title:  Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C
output: pdf_document
geometry: margin = 0.5in
font-size: 11pt
header-includes:
   - \usepackage{caption}
   - \captionsetup[figure]{font=footnotesize, labelfont=bf}
   - \usepackage{sectsty} 
   - \usepackage{helvet}
   - \renewcommand{\familydefault}{\sfdefault}
   - \usepackage{lineno}
   - \linenumbers
   - \renewcommand\linenumberfont{\normalfont\small}
   - \usepackage{setspace}
   - \spacing{1.5}
   - \pagenumbering{arabic}
   - \usepackage{indentfirst}

@@

# Introduction

To form a complete organism, embryonic stem cells must differentiate into a myriad of discrete cellular identities. This is in part accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through histone and DNA modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. Although initially identified for their roles in cellular identity [@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing technologies unexpectedly found many neurodevelopmental disorders (NDDs) are caused by or linked to mutations in chromatin regulators. This relationship is partially explained by their regulation of brain-specific genes or chromatin states, such as modulating genes involved in synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] or the transition between neuronal and glial developmental programs[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also lead to the misexpression of tissue-specific genes outside of their intended environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Currently, very few studies have explored the misexpression of non-neuronal, tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is unclear if this partial loss of brain identity contributes to neurodevelopmental impairments. To elucidate their contribution to neurodevelopmental impairments, it is essential to first characterize the types of genes misexpressed, the developmental time point the dysregulation begins, and the molecular mechanism underlying their de-repression. Characterizing these features will enable us to identify molecular footholds common between NDDs that can then be exploited for potential therapeutics.

In this study, we characterized the misexpression of tissue-enriched genes with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose features include short stature, intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Previous work has demonstrated constitutive _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression and learning impairments[@iwaseMouseModelXlinked2016]. Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus surprisingly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to testis genes, as other tissue-enriched genes have not been systematically evaluated.

Interestingly, some of the ectopic testis transcripts identified in the _Kdm5c_-KO brain are typically expressed in germ cells[@scandagliaLossKdm5cCauses2017]. Unlike somatic cells, germ cells (e.g. sperm and eggs) undergo meiosis and pass on their genetic material to the next generation. The germline and the soma are typically distinguished during early embryogenesis, when germline genes are silenced in epiblast stem cells soon after implantation and only reactivated in a subset to form the germline. Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters. KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in neurons fails to surpress their dysregulation[@scandagliaLossKdm5cCauses2017]. In support of this, KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023]. In support of this, two independent screens in mouse embryonic stem cells (ESCs) recently identified KDM5C as a repressor of _Deleted in azoospermia like (Dazl)_, a key regulator of germline development. However, KDM5C's role in embryonic germline gene repression is currently unclear, given that Dazl is also expressed in ESCs and in the 2-cell stage and germline gene misexpression has yet to be globally characterized during _Kdm5c_-KO embryogenesis.

To elucidate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We identified widespread dysregulation of tissue-enriched genes in both the adult _Kdm5c_-KO brain and EpiLCs, including misexpression of liver, muscle, and ovary genes. The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis-biased genes that are unique to germ cells. To better characterize germline gene misexpression, we then generated a dataset of germline-enriched genes by comparing gene expression in gonads with germ cell depletion. We found _Kdm5c_-KO EpiLCs primarily expressed unique germline genes compared to the mature _Kdm5c_-KO brain, including _Dazl_ and _Stra8_, key drivers of germline identity and meiosis. While KDM5C is directly bound to some germline gene promoters in EpiLCs, it is not directly bound to many germline-enriched mRNAs expressed with _Kdm5c_-KO cells, indicating germline genes can be aberrantly transcribed through indirect mechanisms. Finally, we found KDM5C loss impairs the placement of DNA methylation at germline gene promoters as ESCs differentiate into EpiLCs. Therefore, we propose KDM5C plays a crucial role in the development of tissue identity during early embryogenesis, including establishment of the soma-germline boundary.

**note: need a better conclusion sentence - work on when we know what's happening with last figure/functional consequences**

@@

- **note: should I compare amygdala and hippocampus? scandaglia only looked at hippocampus**

@@

A previous RNA sequencing (RNA-seq) study surprisingly revealed the aberrant expression of testis genes within the adult _Kdm5c_ knockout (-KO) hippocampus[@scandagliaLossKdm5cCauses2017]. Given the high abundance of testis-enriched genes within the mouse transcriptome, it is currently unclear if their misexpression in the _Kdm5c_-KO brain is indicative of random de-repression throughout the genome or instead reflects an impairment in tissue identity. To distinguish these possibilities, we globally assessed the expression of previously characterized tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017] in our previously published mRNA-seq datasets of the _Kdm5c_-KO amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020]

@@

# Discussion
- The demarcation of the germ vs soma is a key feature of multicellularity
- Other H3K4me regulators, anything known about tissue-biased gene expression?
- tissue-biased gene expression:
	- However unlike the gonadal-biased DEGs, many liver and muscle-biased DEGs have a known involvement in brain function. For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory (Figure 1E).
- Otx2 is properly expressed in EpiLCs and prevents pgc identity https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/
